### 2012-09-20 paper_sean/hyperneat.tex
latest change before jeff's read

% !TEX root =main.tex
HyperNEAT is a neuroevolution method which evolves artificial neuron networks (ANNs) using evolutionary algorithms.
HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.

HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
The motivation for CPPNs is that complex patterns can be produced by determining attributes of their phenotypic components as a function of their geometric location. 
This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.

Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop.
For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
These functions allow evolution to exploit various properties such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
Because of evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.

There are links between nodes in CPPNs each of which has a weight used to multiply the signal in each link to determine the magnitude of effect from each node. 
In HyperNEAT, a CPPN genome takes in Cartesian coordinates (X,Y) of the source and the target nodes, and a constant bias value and outputs the weight between the input and hidden layer and a weight between the hidden layer and output (assuming that there is a hidden layer, as in the case for this study). 
All pairwise combinations of source and target nodes are passed into a CPPN to determine the weight of each ANN link.

>>>>

% !TEX root =main.tex
HyperNEAT is a neuroevolution method that evolves artificial neuron networks (ANNs) using evolutionary algorithms.
HyperNEAT has shown to evolve large neural networks that represent many features present in brains such as repetitions and regularities \cite{stanley3}. 
This encoding has previously been applied for gait-learning \cite{yos:clune} and for evolving 2-D and 3-D objects \cite{clune:lipson} and has been successful in displaying complex, natural characteristics such as symmetries and repetition of themes often found in nature.

HyperNEAT indirectly encodes ANNs using compositional pattern producing networks (CPPNs) which represent genomes encoding ANN phenotypes \cite{stanley1}, \cite{stanley2}. 
The motivation for CPPNs is that complex patterns can be produced by determining the attributes of their phenotypic components as a function of their geometric location. 
This is based on the belief that in nature, cells in a body differentiates and specializes in different things with respect to their location. 
In real life, a cell in natural organisms cannot determine its location in space without the help of chemical gradients but \emph{in silico}, cells can be given their geometric coordinates, which CPPNs exploit. 
%A cell in natural organisms cannot determine its location in space by itself and requires chemical gradients are used to signal its location \cite{carroll}. 
%In contrast, \emph{in silico}, cells can be given their geometric coordinates and CPPNs use this to their advantage.

Each CPPN is a directed graph in which every node is a mathematical function, such as sine or Gaussian. 
Moreover, a CPPN genome allows functions to be made of other functions, allowing coordinate frames to be combined and hierarchies to develop, allowing for several different themes to appear such as a repeating theme (sine) with symmetry (Gaussian). 
%For example, a sine function could develop a repeating theme which is passed onto a Gaussian function to create a repeating series of symmetrical motifs. 
These functions allow evolution to exploit various properties that allow HyperNEAT to find regularities without further human intervention \cite{stanley1,clune1}. %such as symmetry (e.g. Gaussian) and repetition (e.g. sine function) \cite{stanley1}. 
%Because evolution is able to exploit such properties, things like gaits are able to feature repetition and symmetry on its own without human intervention \cite{clune1}.

There are weighted links between nodes in CPPNs that are used to multiply the signal in each link and determine the magnitude of effect from each node. 
In HyperNEAT, a CPPN genome takes in Cartesian coordinates (X,Y) of the source and the target nodes, and a constant bias value and outputs the weight between the input and hidden layer and a weight between the hidden layer and output (assuming that there is a hidden layer, as in the case for this study). 
All pairwise combinations of source and target nodes are passed into a CPPN to determine the weight of each ANN link.

### 2012-09-20 paper_sean/methods.tex
latest change before jeff's read

% !TEX root =main.tex
Gaits were first evolved in simulation and then transferred onto the real robot, as previous done in Glette et al. \cite{glette}. 
During the evolution process, we evolved 200 organisms for 200 generations in 20 runs with different initial conditions specified by randomly seeding the run. 
Each gait was 12 seconds long, both in simulation and in reality. 
All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al. \cite{yos:clune} except for the frequency of the sine wave function used, which was changed in order to address the servo shutdown problem mentioned previously. 
We found that gaits that require servos to change their directions often were more prone to servo shutdowns than low frequency gaits. %(at around 1.5Hz per servo). 
But with the sine wave frequency used for Yosinski et al. \cite{yos:clune}, HyperNEAT often evolved high frequency gaits (over 8Hz per servo) which performed poorly in reality due to servo constraints. 
This led to gaits performing dramatically different in real-life compared to simulations. 
%To address this issue, two changes were made:
To address thsi, we reduced by a factor of 6.63 by changing the sine function to

\begin{equation}
sin(2\pi t/250)\pi
\end{equation}

@@

so that HyperNEAT would lean towards evolving low frequency gaits. %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
We further added a policy that punished high frequency gaits while evolving in simulation. 
If the average frequency of each leg for the gait was higher than the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:

@@

After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated for its performance on the real robot.
In real-life trials, the motion of the robot was interpolated linearly between a stationary pose and the commanded gait for two seconds at the beginning and the end. % figure!! 
The position of the robot was measured at the beginning and end of each run, starting from the stationary pose and ending with the stationary pose. 
If a gait performed extremely poorly due to servo shutdowns, the gait was tried again until we could conclude that poor performance is because of the gait and not because of the condition of the robot.

>>>>

% !TEX root =main.tex
Gaits were first evolved in simulation and then transferred onto the real robot, as previous done in Glette et al. \cite{glette}. 
During the evolution process, we evolved 200 organisms for 200 generations in 20 runs with different initial conditions specified by randomly seeding the run. 
Each gait was 12 seconds long, both in simulation and in reality. 
All the HyperNEAT settings used for gait generation was kept the same as in Yosinski et al. \cite{yos:clune} except for the frequency of the sine wave function used, which was changed in order to address the servo shutdown problem mentioned previously after discovering that gaits that require servos to change their directions often were more prone to servo shutdowns than low frequency gaits. %(at around 1.5Hz per servo). 
With the sine wave frequency used for Yosinski et al. \cite{yos:clune} (0.0265Hz), HyperNEAT often evolved high frequency gaits (over 8Hz per servo) that performed poorly in reality due to servo constraints. 
This led to dramatically different gait performance in real-life compared to simulations. 
To address this, we changed the frequency of sine function to 0.004Hz so that HyperNEAT would lean towards evolving low frequency gaits.

%\begin{equation}
%sin(2\pi t/250)\pi
%\end{equation}

@@

 %Sine frequency of 0.004 was used because we experimentally found that HyperNEAT was able to produce natural and high-performing gaits with low frequencies with this frequency. 
We also added a policy that punished high frequency gaits while evolving in simulation. 
If the average frequency of each leg for the gait was higher than the threshold frequency, the fitness was punished by exponentially reducing the distance travelled by the robot:

@@

After evolving the gaits in simulation, the champion gait of the last generation of each of the 20 runs was transferred onto the real robot and evaluated for its performance on the real robot.
In real-life trials, the motion of the robot was interpolated linearly between a stationary pose and the commanded gait for two seconds at the beginning and the end. 
The position of the robot was measured at the beginning and the end of each run, starting from the stationary pose and ending with the stationary pose. 
If a gait performed extremely poorly due to servo shutdowns, the gait was tried again until we could conclude that poor performance is because of the gait and not due to the condition of the robot such as overheated servos.

### 2012-09-20 paper_sean/robot.tex
latest change before jeff's read

The body of the robot was printed using an Objet Connex 500 3-D Printing System and actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
The servos were attached to the robot using off-the-shelf components and parts.
The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree].

The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} that would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}. 
The ones most relevant to this study were shutdowns due to high torque and range. 
As has been reported previously with QuadraTot \cite{yos:clune}, many of the gaits that were tested on the QuadraTot forced the servos to exert too much torque or would command out of range positions and cause the servos to shutdown. In order to address this problem and avoid collisions with its own body, a motion cropping system was implemented on the QuadraTot. 
In addition to placing limits on the range of motions for the inner joints, outer joints, and the hip joint of [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] respectively, an extra cropping system named the smart cropping system was adopted from Shen et al. \cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque. %This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC/SL - do they sum the position values, or the joint angles? Since all the previous extremes are expressed in angles, the writing makes me expect angles... / They do use the position values and I think that it would be best to be consistent with what Jason/Haocheng did. This is also the way g is calculated in the simulator and the python drivers as well and more accurately reflects what we actually did. 
%and modifying the servo positions from ``fierce gaits'', which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$. %JMC/SL - the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places. No need to invent a definition only to use it 2-3 times...can you swap in different, more descriptive words? / Again, I am trying to be consistent with what Jason and Haocheng used. I agree that 'fierce' is not the best word but if it is different from what J/H did, then there are inconsistencies between the publications...
%We set the minimum threshold \emph{\textbf{g}} to be 730, found through experimentation to provide a good balance between performance and safety. %JMC/SL - This description is not great. The last sentence's use of g doesn't make sense with the equation...is g a threshold or the sum? Please try to clean up this description. / Think this clarified things

The robot was controlled by CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10. 
In a previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.
The servos were sent positions from the computer at 40Hz via the Pydynamixel library.

We identified the robot's position via infrared LEDs on the robot that could be tracked by a Nintendo Wii remote.
%JMC/SL you sure there was more than one IR LED? I thought it was just one IR and one visible for us to tell if there is power to the LEDs. If it is just one, switch to singular instead of plural in the previous sentence / Yep there are like 3 IR LEDs being used.
To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library \cite{yos:clune}. 
%JMC/SL The paragraph about the Wii mote was very wordy. I think I trimmed it by half. You should try to be similarly compact the rest of the paper. Compare your version to mine and try to see how I condensed things. / Will do boss.

>>>>

The body of the robot was printed using an Objet Connex 500 3-D Printing System and is actuated using 9 Robotis Dynamixel servos: 5 AX-18A servos and 4 AX-12A servos. 
The stronger AX-18A servos were used to actuate the inner joints of the robot and the hip joint while the AX-12A servos were used to actuate the outer joints. 
The servos have a range of motion of [0, 1023], roughly equivalent to a physical range of [-12\degree, +120\degree].

The Dynamixel AX-12A and AX-18A servos have a safety feature called \emph{Alarm Shutdown} that would turn off the servo if the servo's current, range, temperature, or torque was too high \cite{robotis}. 
The ones most relevant to this study were shutdowns due to high torque and range. 
It has been reported previously Yosinski et al. \cite{yos:clune}, that many gait tested on the QuadraTot forced the servos to exert too much torque or command positions that are out of range and cause the servos to shutdown. In order to address this problem and avoid collisions with its own body, we implemented a motion cropping system by limiting the possible range of movement of the QuadraTot by [-85\degree, +60\degree], [-113\degree, +39\degree], and [-28\degree, +28\degree] for the inner, outer, and hip joints respectively.
In addition to this, we implemented an extra cropping system named the Smart Cropping System from Shen et al. \cite{haocheng} to further prevent servo shutdowns from leg positions where the combination of the two angles interact to produce an extreme amount of torque. %This smart cropping algorithm calculates a leg position by summing the servo position values from the inner and outer joints of a leg, \emph{\textbf{g}} (\emph{\textbf{g}} = $j_{inner}$ + $j_{outer}$) %JMC/SL - do they sum the position values, or the joint angles? Since all the previous extremes are expressed in angles, the writing makes me expect angles... / They do use the position values and I think that it would be best to be consistent with what Jason/Haocheng did. This is also the way g is calculated in the simulator and the python drivers as well and more accurately reflects what we actually did. 
%and modifying the servo positions from ``fierce gaits'', which usually have relatively low \emph{\textbf{g}}s, to remap them onto the boundary of a quasi-trangular area in a 2-D space defined by $j_{inner}$ and $j_{outer}$. %JMC/SL - the term fierce sounds silly and non-professional (sorry!) I took out the place where you defined it, and we only use it in two other places. No need to invent a definition only to use it 2-3 times...can you swap in different, more descriptive words? / Again, I am trying to be consistent with what Jason and Haocheng used. I agree that 'fierce' is not the best word but if it is different from what J/H did, then there are inconsistencies between the publications...
%We set the minimum threshold \emph{\textbf{g}} to be 730, found through experimentation to provide a good balance between performance and safety. %JMC/SL - This description is not great. The last sentence's use of g doesn't make sense with the equation...is g a threshold or the sum? Please try to clean up this description. / Think this clarified things

We controlled the robot using a CompuLab Fit-PC2, a compact computer running Ubuntu Linux 10.10 and commanded the servos at 40Hz via the Pydynamixel library.

%In a previous study \cite{yos:clune}, this computer was placed on the robot to keep the robot untethered, but for this study the computer was taken off the robot to reduce weight, because the weight of the robot seemed to be one of the factors contributing to servo shutdowns.

We identified the robot's position using infrared LEDs on the robot which was tracked using a Wiimote as in Yosinski et al. \cite{yos:clune}.
%JMC/SL you sure there was more than one IR LED? I thought it was just one IR and one visible for us to tell if there is power to the LEDs. If it is just one, switch to singular instead of plural in the previous sentence / Yep there are like 3 IR LEDs being used.
%To maximize its viewable window, the Wii remote was fixed to the ceiling pointing down. It communicated with the robot computer via bluetooth using the CWiid library \cite{yos:clune}. 
%JMC/SL The paragraph about the Wii mote was very wordy. I think I trimmed it by half. You should try to be similarly compact the rest of the paper. Compare your version to mine and try to see how I condensed things. / Will do boss.

### 2012-09-20 paper_sean/simulator.tex
latest change before jeff's read

% !TEX root =main.tex
Simulations occurred in C++ software customized for the QuadraTot\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} in the NVIDIA PhysX physics simulator, as described previously \cite{glette}. 
Specifically, the simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components. 
%The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
%The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.%JMC: Assumed
%The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. %JMC: Assumed
%The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. %JMC: Assumed
%In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
In an attempt to more accurately model the robot behavior, we implemented the same cropping function in the simulator as in the actual robot. %JMC/SL - Do all results in this paper reflect this? / They will soon.
Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
The friction values used for this study was the same as that used in Glette et al. \cite{glette}. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%

>>>>

% !TEX root =main.tex
The simulator\footnote{Available for Linux and Windows at http://quadratot.yosinski.com} used for this study was the same as the one used in Glette et al. \cite{glette}. The simulator is written in C++ using the Nvidia PhysX engine. The simulation captured the number and type of joints, mass, and rough shapes and lengths of the QuadraTot components to model the robot. Variables such as the friction coefficient were kept the same for this study. 
%Simulations occurred in C++ software customized for the in the NVIDIA PhysX physics simulator, as described previously \cite{glette}. 
%The simulator was visualized by OpenGL library to allow the user to have a more intuitive grasp of what is going on. %JMC Unnecessary detail since this paper is not about the simulator. 
%The simulator visualization included a recording feature which allows the user to create videos of the robot in simulation.%JMC: Assumed
%The simulator also features a command line function which allows the users to input a gait file and run it on the simulator. %JMC: Assumed
%The simulator then outputs a file containing the X, Y, Z coordinates of the center of mass of the robot over time. %JMC: Assumed
%In simulation, the robot traverses in the XY plane with the Z axis representing the height off the ground. %JMC Unnecessary detail
We implemented the initial cropping function in the simulator as in the actual robot but not the smart cropping function because with the smart cropping function implemented in simulation, there was too much limitation in the movements which drastically limited the possible gaits HyperNEAT could explore. %JMC/SL - Do all results in this paper reflect this? / They will soon.
%Although the simulator was originally written in Windows for the Windows operating platform we ported and ran it on Linux (Ubuntu 11.04) to seamlessly link the simulator and the already exising libraries together. 
%
% and uses OpenGL and GLFW libraries for visualization. The simula- tor was originally written for the Windows operating system but was ported and run on Linux (Ubuntu 11.04) for this ex- periment. It features a commandline function which allows the user to specify a gait input file and extract an output file with the X, Y, Z coordinate information over time. The Eu- clidean distance travelled by the robot in the XY plane was used to calculate the speed of the robot in simulation and was used as the fitness value used for evolution.
%
